|

### 2022-01-26 submission/manuscript.rmd
add treatment interaction scenarios

---
title: |
  An interaction between baseline risk and treatment assignment 
  generally imporoved treatment effect predictions
abstract: |
 **Objective**: To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting**: We simulated RCT data using diverse assumptions for the
 average treatment effect, a baseline prognostic index of risk (PI),
 the shape of its interaction with treatment (none, linear, quadratic
 or non-monotonic) and the magnitude of treatment-related harms. In
 each sample we predicted absolute benefit using: models with a
 constant relative treatment effect; stratification in quarters of the
 PI; models including a linear interaction of treatment with the PI;
 models including an interaction of treatment with a restricted cubic
 spline (RCS) transformation of the PI; an adaptive approach using
 Akaike’s Information Criterion. We evaluated predictive performance
 using root mean squared error and measures of discrimination and
 calibration for benefit. **Results**: The linear-interaction model
 and the RCS-interaction model outperformed the constant treatment
 effect model in many simulation scenarios. The RCS-model was optimal
 when quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. **Conclusion**: An
 interaction between risk and treatment assignment generally improved
 treatment effect predictions. Non-linear RCS interactions should be
 considered for larger sample size.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

# Introduction
Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
at the development of models predicting either individualized effects or which
of two (or more) treatments is better for an individual [@Varadhan2013]. In
prior work, we divided such methods in three broader categories based on the
reference class used for defining patient similarity when making individualized
predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
prediction of baseline risk as the reference; treatment effect modeling
approaches also model treatment-covariate interactions, in addition to risk
factors; optimal treatment regime approaches focus on developing treatment
assignment rules and therefore rely heavily on modeling treatment effect
modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. This two-step approach identified substantial absolute treatment
effect differences between low-risk and high-risk patients in a re-analysis of
32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
level may be accurate, these estimates do not apply to individual patients,
especially for patients with predicted risk at the boundaries of the risk
intervals. Hence, the risk-stratified approach is useful for exploring and
presenting HTE, but is not useful for supporting treatment decisions for
individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
different risk-based models for predicting individualized treatment effects. We
simulated RCT settings to compare the performance of these models under
different assumptions of the relationship between baseline risk and treatment.
We illustrated the different models by a case study of predicting individualized
effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
patients with an acute myocardial infarction (MI).

@@

We simulated a typical RCT that is undertaken to compare a binary outcome
(e.g. death) between a group of treated patients in the treatment arm and a
group of untreated patients in the control arm. For each patient we generated 8
baseline covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1,
0.2)$. Treatment was allocated using a 50:50 split. Outcomes for patients in the
control arm were generated from a logistic regression model including all
baseline covariates. In the base scenarios coefficient values were such, that
the AUC of the logistic regression model was $0.75$ and the event rate in the
control arm was $20\%$. Binary outcomes in the control arm were generated from
Bernoulli variables with true probabilities
$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.

@@

The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
respectively. We also evaluated the effect of worse or better discriminative
ability for risk, adjusting the baseline covariate coefficients, such that the
AUC of the regression model in the control arm was 0.65 and 0.85 respectively.

@@

A *stratified HTE method* has been suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute treatment effects within risk strata
are estimated by the difference in event rate between patients in the control
arm and patients in the treated arm. We considered this approach as a reference,
expecting it to perform worse than the other candidates, as its objective is not
to individualize benefit prediction.

@@

Finally, we considered an adaptive approach using Akaike's Information Criterion
(AIC) for model selection. The candidate models were: a constant treatment
effect model, a model with a linear interaction with treatment and RCS models
with 3, 4 and 5 knots.

@@

The linear interaction model outperformed all RCS methods in terms of
RMSE in scenarios with true constant relative treatment effect (OR =
0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
quadratic deviations from a constant relative treatment effect (Figure
\ref{fig:rmsebase}; panels A-C). However, with non-monotonic
deviations from a constant relative treatment effect, the RMSE of the
linear interaction model increased substantially, especially in the
presence of treatment-related harms (Figure \ref{fig:rmsebase}; panel
D). In these scenarios, RCS-3 outperformed all other methods in terms
of RMSE. The constant treatment effect approach had overall best
performance under true constant treatment effect settings, but was
sensitive to all considered deviations, resulting in increased
RMSE. Finally, the adaptive approach had comparable performance to the
best-performing method in each scenario. However, in comparison with
the best-performing approach, its RMSE was more variable across the
500 replications in the scenarios with linear and non-monotonic
deviations, especially when also including moderate or strong
treatment-related harms. This is caused by wrongly selecting the
constant treatment effect model in a substantial proportion of the
replications (Supplement, Figure S1).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

Increasing the sample size to `r 17000` favored RCS-3 the most, as it
achieved lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower RMSE (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of the strength of treatment-related
harms. Due to the large sample size, the RMSE of the adaptive approach
was even more similar to the best-performing method, and the constant
relative treatment effect model was less often wrongly selected
(Supplement, Figure S2).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

@@

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

In comparison with the true approach, dsicrimination for benefit in
the scenario with a constant relative treatment effect was only
slightly lower for the linear interaction model, but substantially
lower for the non-linear RCS approaches (Figure
\ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods
discriminated quite similarly (Figure \ref{fig:discrimination}; panels
B-C). In the scenario with non-monotonic deviations, the constant
effect model had much lower discriminative ability compared to
all other methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel
D). The adaptive approach was unstable in terms of discrimination for
benefit, especially in the presence of treamtnt-related harms. With
increasing number of RCS knots, we observed decreasing median values
and increasing variability of the c-for-benefit in all scenarios. When
we increased the sample size to 17,000 we observed similar trends,
however the performance of all methods was more stable (Supplement,
Figure S4). Finally, when we increased the true prediction AUC to 0.85
the adaptive approach in the case of non-monotonic deviations was,
again, quite unstable, especially with null or moderate
treatment-related harms (Supplement, Figure S5). In these scenarios,
the adaptive approach tended to select more often the much inferior
constant treatment effect method (Supplement, Figure S3)

```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

treatmentArms <- gusto %>%
  group_by(tpa) %>%
  summarise(n = n())
```

We demonstrate the different methods for individualizing treatment
benefits using data from `r nrow(gusto)` patients with an acute
myocardial infarction (MI) included in the GUSTO-I trial.
`r treatmentArms %>% filter(tpa == 1) %>% select(n)` patients were randomized to
tissue plasminogen activator (tPA) treatment and
`r treatmentArms %>% filter(tpa == 0) %>% select(n)` were randomized to
streptokinase. The outcome of interest was 30-day mortality, recorded
for all patients.

@@

We used the risk linear predictor to fit the proposed methods under
study for individualizing absolute benefit predictions. All methods
predicted increasing benefits for patients with higher baseline risk
predictions, but the fitted patterns were clearly different. The
adaptive approach selected the model with RCS smoothing with 4
knots. However, for very low baseline risk the decreasing predicted
benefit with increasing risk may be somewhat too flexible. The more
robust models, the linear interaction model or the model with RCS
smoothing (3 knots), gave very similar benefit predictions, followed
the evolution of the stratified estimates very closely and may
therefore be preferable for use in clinical practice. The linear
interaction model had somewhat lower AIC compared to the model with
RCS smoothing (3 knots), slightly better cross-validated
discrimination (c-for-benefit
`r round(discriminationPerformance[2, ], 3)` vs
`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
`r round(calibrationPerformance[3, ], 4)`).

@@

RCS-4 and RCS-5 proved to be too flexible in all considered scenarios,
as indicated by higher RMSE, increased variability of discrimination
for benefit and worse calibration of benefit predictions. Even with
larger sample sizes and strong quadratic or non-monotonic deviations
from the base case scenario of constant relative treatment effects,
these more flexible restricted cubic splines did not outperform the
simpler RCS-3 These approaches may only be helpful if we expect more
extreme patterns of heterogeneous treatment effects compared to the
quadratic deviations considered here.

The constant treatment effect model, despite having adequate
performance in the presence of weak treatment effect heterogeneity on
the relative scale, quickly broke down with stronger deviations from
constant relative treatment effects. In these cases, the stratified
approach generally had lower error rates compared to the constant
treatment effect model. Stepwise treatment benefit estimates are very
useful for demonstrating treatment effect heterogeneity–because
estimating treatment effect requires groups of patients rather than
individual patients–but are not helpful for making individualized
absolute benefit predictions.

Increasing the discriminative ability of the risk model–by increasing
the predictor coefficients of the true risk model–reduced RMSE for all
methods. This increase in discriminative ability translates in higher
variability of predicted risks, which, in turn, allows the considered
methods to better capture absolute treatment benefits. As a
consequence, the increase in discriminative ability of the risk model
also led to higher values of c-for-benefit. Even though risk model
performance is very important for the ability of risk-based methods to
predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

The adaptive approach had adequate performance, following closely on
average the performance of the “true” model in most
scenarios. However, with smaller sample sizes it tended to “miss” the
treatment-risk interactions and selected simpler models (Supplementary
Table S7). This resulted in increased RMSE variability in these
scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in
the case of smaller sample sizes the simpler linear interaction model
is a safer choice for predicting absolute benefits.

Risk-based approaches to predictive HTE estimate treatment benefit as
a function of baseline risk. A limitation of our study is that we
assumed treatment benefit to be a function of baseline risk in the
majority of the simulation scenarios. We attempted to address that by
introducing constant moderate and strong treatment-related harms,
applied on the absolute scale. Also, we considered a small number of
scenarios with true treatment-covariate interactions, in which our
main conclusions remained the same (Supplement, XX). Future simulation
studies could explore the effect of more extensive deviations from
risk-based treatment effects.

@@

In conclusion, the linear interaction approach is a viable option with
smaller sample sizes and/or moderately performing risk prediction
modelsRCS-3 is a better option when non-monotonic deviations from a
constant relative treatment effect and/or substantial
treatment-related harms are anticipated. With larger sample size, an
adaptive approach -- selecting the method with optimal AIC -- can also
be considered as an automated alternative.

>>>>

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal
 prediction of individualized treatment effects from RCTs. **Study
 Design and Setting**: We simulated RCT data using diverse assumptions
 for the average treatment effect, a baseline prognostic index of risk
 (PI), the shape of its interaction with treatment (none, linear,
 quadratic or non-monotonic) and the magnitude of treatment-related
 harms (none or constant independent of the PI). In each sample we
 predicted absolute benefit using: models with a constant relative
 treatment effect; stratification in quarters of the PI; models
 including a linear interaction of treatment with the PI; models
 including an interaction of treatment with a restricted cubic spline
 (RCS) transformation of the PI; an adaptive approach using Akaike’s
 Information Criterion. We evaluated predictive performance using root
 mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the
 RCS-interaction model outperformed the constant treatment effect
 model in many simulation scenarios. The RCS-model was optimal when
 quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. Illustrations in the
 GUSTO-I trial confirmed these findings. **Conclusion**: An
 interaction between baseline risk and treatment assignment should be
 considered to improve treatment effect predictions. Non-linear RCS
 interactions should be considered only in larger sample sizes.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

# Introduction
Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
at the development of models predicting either individualized effects or which
of two (or more) treatments is better for an individual [@Varadhan2013]. In
prior work, we divided such methods in three broader categories based on the
reference class used for defining patient similarity when making individualized
predictions or recommendations [@Rekkas2020]. First, risk-modeling approaches
use prediction of baseline risk as the reference; second, treatment effect
modeling approaches also model treatment-covariate interactions, in addition to
risk factors; third, optimal treatment regime approaches focus on developing
treatment assignment rules and rely heavily on modeling treatment effect
modifiers. A key difference between these approaches is their parsimony in
dealing the treatment effect modifiers, with no interaction considered (risk
modeling), a limited number of interactions (effect modeling), or a larger set
of interactions (optimal treatment regime approaches).

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. This two-step approach identified substantial absolute treatment
effect differences between low-risk and high-risk patients in a re-analysis of
32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
level may be accurate, these estimates may need further refinement for
individual patients, especially for patients with predicted risk at the
boundaries of the risk intervals. Hence, the risk-stratified approach is useful
for exploring and presenting HTE, but is not sufficient for supporting treatment
decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
different risk-based models for predicting individualized treatment effects. We
simulated RCT settings to compare the performance of these models under
different assumptions of the relation between baseline risk and treatment. We
illustrated the different models by a case study of predicting individualized
effects of treatments for acute myocardial infarction (MI) in a large randomized
controlled trial (RCT).

@@

We simulated a typical RCT that is undertaken to compare a binary outcome
(e.g. death) between a group of patients in the treatment arm and a
group of untreated patients in the control arm. For each patient we generated 8
baseline covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1,
0.2)$. Treatment was allocated using a 50:50 split. Outcomes for patients in the
control arm were generated from a logistic regression model including all
baseline covariates. In the base scenarios coefficient values were such, that
the AUC of the logistic regression model was $0.75$ and the event rate in the
control arm was $20\%$. Binary outcomes in the control arm were generated from
Bernoulli variables with true probabilities
$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.

@@

The sample size for the base scenarios was set to 4,250, since this sample size
provides $80\%$ power for the detection of a marginal OR of 0.8 with the
standard alpha of 0.5\%. We evaluated the effect of smaller or larger sample
sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
respectively. We also evaluated the effect of worse or better discriminative
ability for risk, adjusting the baseline covariate coefficients, such that the
AUC of the regression model in the control arm was 0.65 and 0.85 respectively.

@@

A *stratified HTE method* has been suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute treatment effects within risk strata
are estimated by the difference in event rate between patients in the control
arm and patients in the treated arm. We considered this approach as a reference,
expecting it to perform worse than the other candidates, as its objective is to
provide an illustration of HTE rather than to optimize individualized benefit
predictions.

@@

Finally, we considered an adaptive approach using Akaike's Information Criterion
(AIC) for model selection. The candidate models were: a constant treatment
effect model, a model with a linear interaction with treatment and RCS models
with 3, 4 and 5 knots. The extra degrees of freedom were 1 (linear interaction),
2, 3 and 4 (RCS models) for these increasingly complex interactions with the
treatment effect.

@@

The linear interaction model outperformed all RCS methods in terms of RMSE in
scenarios with true constant relative treatment effect (OR = 0.8, N = `r 4250`
and AUC = 0.75), strong linear and even strong quadratic deviations from a
constant relative treatment effect (Figure \ref{fig:rmsebase}; panels
A-C). However, with non-monotonic deviations from a constant relative treatment
effect, the RMSE of the linear interaction model increased substantially,
especially in the presence of treatment-related harms (Figure
\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
methods in terms of RMSE. As might be expected the constant treatment effect
approach had overall best performance under true constant treatment effect
settings. It was sensitive to all considered deviations, resulting in increased
RMSE. Finally, the adaptive approach had comparable performance to the
best-performing method in each scenario. However, in comparison with the
best-performing approach, its RMSE was more variable in the scenarios with
linear and non-monotonic deviations, especially when also including moderate or
strong treatment-related harms. On closer inspection, we found that this
behavior was caused by wrongly selecting the constant
treatment effect model in a substantial proportion of the replications
(Supplement, Figure S1). This problematic behavior was less with larger sample
sizes (see below).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

Increasing the sample size to `r 17000` favored RCS-3 the most, It achieved
lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower RMSE (median
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of the strength of treatment-related
harms. Due to the large sample size, the RMSE of the adaptive approach
was even more similar to the best-performing method, and the constant
relative treatment effect model was less often wrongly selected
(Supplement, Figure S2).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

@@

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250. True prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

In comparison with the true approach, dsicrimination for benefit in the scenario
with a constant relative treatment effect was only slightly lower for the linear
interaction model, but substantially lower for the non-linear RCS approaches
(Figure \ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods discriminated
quite similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario
with non-monotonic deviations, the constant effect model had much lower
discriminative ability compared to all other methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel D). The
adaptive approach was unstable in terms of discrimination for benefit,
especially in the presence of treatment-related harms. With increasing number of
RCS knots, we observed decreasing median values and increasing variability of
the c-for-benefit in all scenarios. When we increased the sample size to 17,000
we observed similar trends, however the performance of all methods was more
stable (Supplement, Figure S4). Finally, when we increased the true prediction
AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,
again, quite unstable, especially with null or moderate treatment-related harms
(Supplement, Figure S5). In these scenarios, the adaptive approach tended to
select more often the inferior constant treatment effect method (Supplement,
Figure S3)

```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250. True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

gusto <- gusto %>%
  tibble() %>%
  filter(tx != "SK+tPA") %>%
  rename(
    "outcome" = "day30",
    "treatment" = "tpa"
  )

treatmentArms <- gusto %>%
  group_by(treatment) %>%
  summarise(n = n())
```

We demonstrate the different methods for individualizing treatment benefits
using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
included in the GUSTO-I trial.
`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
patients were randomized to tissue plasminogen activator (tPA) treatment and
`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
were randomized to streptokinase. The outcome of interest was 30-day mortality
(total of `r sum(gusto$outcome)` events), recorded for all patients.

@@

We used the risk linear predictor to fit the proposed methods under study for
individualizing absolute benefit predictions. All methods predicted increasing
benefits for patients with higher baseline risk predictions, but the fitted
patterns were somewhat different. The adaptive approach selected the model with
RCS smoothing with 4 knots. However, for very low baseline risk the decreasing
predicted benefit with increasing risk may be somewhat too flexible. The more
robust models, such as the linear interaction model and the model with RCS
smoothing (3 knots), gave very similar benefit predictions. These models
followed the evolution of the stratified estimates very closely and may
therefore be preferable for use in clinical practice. The linear interaction
model had somewhat lower AIC compared to the model with RCS smoothing (3 knots),
slightly better cross-validated discrimination (c-for-benefit
`r round(discriminationPerformance[2, ], 3)` vs
`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
`r round(calibrationPerformance[3, ], 4)`).

@@

RCS-4 and RCS-5 proved to be too flexible in all considered scenarios, as
indicated by higher RMSE, increased variability of discrimination for benefit
and worse calibration of benefit predictions. Even with larger sample sizes and
strong quadratic or non-monotonic deviations from the base case scenario of
constant relative treatment effects, these more flexible restricted cubic
splines did not outperform the simpler RCS-3 These approaches may only be
helpful if we expect more extreme patterns of heterogeneous treatment effects
compared to the quadratic deviations considered here. Considering interactions
in RCS-3 models as the most complex approach often may be reasonable.

The constant treatment effect model, despite having adequate performance in the
presence of weak treatment effect heterogeneity on the relative scale, quickly
broke down with stronger deviations from constant relative treatment effects. In
these cases, the stratified approach generally had lower error rates compared to
the constant treatment effect model. Such stepwise treatment benefit estimates
are useful for visually demonstrating treatment effect heterogeneity but may be
considered insufficient for making individualized benefit predictions.

Increasing the discriminative ability of the risk model–by increasing the
predictor coefficients of the true risk model–reduced RMSE for all methods. This
increase in discriminative ability translates in higher variability of predicted
risks, which, in turn, allows the considered methods to better capture absolute
treatment benefits. As a consequence, the increase in discriminative ability of
the risk model also led to higher discrimination between those with low or high
benefit (as reflected in values of c-for-benefit). Even though risk model
performance is very important for the ability of risk-based methods to predict
treatment benefit, prediction model development was outside the scope of this
work and has already been studied extensively [@vanKlaveren2019; @Burke2014;
@Abadie2018].

The adaptive approach had adequate performance, following closely on average the
performance of the “true” model in most scenarios. With smaller sample sizes it
tended to miss the treatment-risk interactions and selected simpler models
(Supplementary Table S7). This conservative behavior resulted in increased RMSE
variability in these scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in the case of
smaller sample sizes the simpler linear interaction model may be a safer choice
for predicting absolute benefits, if a non-constant treatment effect is
suspected.

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work (REF!!). Also, we considered a small number of scenarios with
true treatment-covariate interactions, in which our main conclusions remained
the same (Supplement, XX). Future simulation studies could explore the effect of
more extensive deviations from risk-based treatment effects.

@@

In conclusion, the linear interaction approach is a viable option with smaller
sample sizes and/or moderately performing risk prediction models if we consider
a non-constant relative treatment effect plausible. RCS-3 is a better option
when non-monotonic deviations from a constant relative treatment effect and/or
substantial treatment-related harms are anticipated. With larger sample size, an
adaptive approach -- selecting the method with optimal AIC -- can also be
considered as an automated alternative.
